---
title: 'Activity 4: Tidyverse Data Wrangling'
author: "Phoebe Arriola"
date: "2026-06-01"
output: html_document
---
# Tidyverse Wrangling Operations

## Loading Packages and Initial Inspection
```{r load-tidyverse, message = FALSE, warning = FALSE}
library(tidyverse)
library(palmerpenguins)

# Peek at the integrated iris data structure using tidyverse pipe
iris |> glimpse()
```

## Data Transformation Using Dplyr Verbs
Demonstrating filtering, column picking, and mutations.

### Filtering and Selecting Rows/Columns
```{r filtering-selections}
# Filter rows matching setosa
filter(iris, Species == "setosa")

# Select targeted metrics columns
select(iris, Sepal.Length, Sepal.Width, Species)

# Inverted column matches (Removing Sepal columns)
select(iris, -contains("Sepal"))
```

### Piping Pipelines (`|>` or `%>%`) and Mutating Data
```{r mutations}
# Generating ratios and arranging them
sepal.sort <- iris |>  
  mutate(Sep.LWratio = Sepal.Length/Sepal.Width) |> 
  arrange(Sep.LWratio)

# Use knitr's kable function to output a clean table snippet
knitr::kable(head(sepal.sort), digits = 2)
```

## Advanced Aggregations and Pivoting Operations

### Grouping and Summarizing Penguin Metrics
```{r penguin-summary}
penguins |> 
  filter(!is.na(sex)) |>
  group_by(species, sex) |> 
  summarize(mean_mass = mean(body_mass_g)) 
```

### Toggling Data Formats (Wide to Long)
```{r pivoting}
# Construct original manual broad dataframe
set.seed(1)
plots <- paste0("plot", LETTERS[1:4])
sunflower <- rpois(n=4, 20)
aster <- rpois(n=4, 50)
dandelion <- rpois(n=4, 80) 

Flowers <- data.frame(plots, sunflower, aster, dandelion)

# Transform wide layout to long layout using pivot_longer
FlowersLong <- Flowers |> pivot_longer(cols = !plots,
                        names_to = "flowers",
                        values_to = "counts")

# Show converted long framework using kable
knitr::kable(FlowersLong)
```