---
title: "Activity 07: K-means Clustering"
author: "Phoebe Arriola"
date: "2026-06-08"
output: html_document
---
```{r part-a}
####
## A. K-means (Single K)
####

# 1. Read in the morphology data as a dataframe
getwd()
setwd("/Users/phoebearriola/Desktop/uni")
list.files()
morph_data <- read.csv("morphology_data.csv")

# 2. Remove the Sex column 
numeric_data <- morph_data[, !names(morph_data) %in% c("Pop", "Sex")]

# 3. Ensure that the dataframe has no NAs
good <- complete.cases(numeric_data)
numeric_data <- numeric_data[good, ]

# 4. Standardize all values using the "scale" function
morph_scaled <- scale(numeric_data)

# 5. Set seed for reproducibility
set.seed(123)

# 6. Perform K-means clustering with 3 clusters and 25 initial random centroids
kmeans_result3 <- kmeans(morph_scaled, centers = 3, nstart = 25)

# 7. Show the cluster assignment and a summary of results
head(kmeans_result3$cluster)
kmeans_result3

# 8. Create a scatter plot with points colored by the assigned cluster.
plot(
  morph_scaled[, 1:2],
  col = kmeans_result3$cluster,
  pch = 19,
  xlab = "SVL (scaled)",
  ylab = "TD (scaled)",
  main = "K-means Clustering of Lizard Morphology Data (K = 3)"
)

# adding the centroids to the plot 
points(
  kmeans_result3$centers[, 1:2],
  col = 1:3,
  pch = 8,
  cex = 2,
  lwd = 2
)


```{r part-b}
####
## B. K-means (loop)
####

# 1. Create a loop function to perform K-means with cluster centers from 2 to 10
wss_values <- numeric(10)
for (k in 2:10) {
  wss_values[k] <- kmeans(morph_scaled, centers = k, nstart = 25)$tot.withinss
}

# 2. From the output of your loop function, create an elbow curve plot to determine optimal K
plot(
  2:10, 
  wss_values [2:10], 
  type = "b", 
  pch = 19, 
  xlab = "Number of Clusters K",
  ylab = "Total Within-Cluster Sum of Squares (WSS)",
  main = "Elbow Method Curve via Loop Function"
)

# 3. Recreate steps 6–8 using the optimal K determined by the elbow curve
# in the dataset, there is a distinct bend in K = 2
optimal_k <- 2 
set.seed(123)
kmeans_optimal <- kmeans(morph_scaled, centers = optimal_k, nstart = 25)

# to see the optimal cluster assignments and summary results
head(kmeans_optimal$cluster)
print(kmeans_optimal)

# creating a final scatter plot with points colored by the optimal clusters
plot(
  morph_scaled[, 1:2],
  col = kmeans_optimal$cluster,
  pch = 19,
  xlab = "SVL (scaled)",
  ylab = "TD (scaled)",
  main = paste("K-means Clustering with Optimal K =", optimal_k)
)

# adding the final optimal centroids
points(
  kmeans_optimal$centers[, 1:2],
  col = 1:optimal_k,
  pch = 8,
  cex = 2,
  lwd = 2
)